---
title: "HW 4 - Point Pattern Analysis"
date: "Due Tuesday, Sept. 24th, 2024 via GitHub"
output: pdf_document
geometry: top=1in, bottom=1in, left=1in, right=1in
---

***  
```{r Libraries}
library(terra)   
library(ggplot2) 
library(viridis)
library(raster)
library(ncf)
library(sp)
library(gstat)
library(sf)
library(spatstat)
library(readr)
library(dplyr)
```

Assignment: This assignment involves analyses of both simulated and real point-pattern datasets. As in
previous coding assignments, you will be graded on your answers and your ability to produce clean, well
commented R code that performs the tasks listed below. If you used AI code generating tools, please indicate
this in your submission.
Write R scripts to complete the following tasks and answer each question.
1. Simulate three types of point-patterns: (i) Complete Spatial Randomness (CSR), (ii) clustered, and (iii)
segregated. Convert each to a ppp object and plot your simulated point-patterns (be sure to add an
appropriate title so I know which is which).
For the point-pattern with CSR: what is the value of lambda and what is the definition of this parameter?
```{r Question 1}
# Make a window
win <- owin(c(0, 1), c(0, 1)) # A unit square

# CSR
lambda_csr <- 100 #intensity 
ppp_csr <- rpoispp(lambda_csr, win=win)

# Clustered Pattern
lambda_cluster <- 5  #still intensity
r_cluster <- 0.05     #size or radius of cluster
k_cluster <- 10        # points/cluster
ppp_cluster <- rMatClust(lambda_cluster, r_cluster, k_cluster, win=win)

# (iii) Segregated Pattern 
r_inhibit <- 0.05 
ppp_segregated <- rSSI(r_inhibit, n=50, win=win)

#Plot stuff
par(mfrow=c(1, 3))  # 1 row, three plots

# CSR plotn
plot(ppp_csr, main="(CSR)")

# Clustered pattern
plot(ppp_cluster, main="Clustered Pattern")

# Segregated pattern
plot(ppp_segregated, main="Segregated (Inhibitory) Pattern")


mtext("Lambda is 100, and it represents the number, on average,", 
      side = 1, line = -3, adj = 1, col = "magenta", font = 2, cex = 1.5)
mtext("of observations within the defined square", 
      side = 1, line = -1, adj = 1, col = "magenta", font = 2, cex = 1.5)

```

2. Analyze each point pattern using G-, K- and F-tests and assess the significance of each test using null
models and a Monte Carlo approach. Provide a brief interpretation of each of your simulated patterns
as informed by the G-, K- and F-tests.
```{r Question 2, pat 1: CSR}
# CSR: Complete Spatial Randomness First


# Test, G- F-, and K- stuff
g_csr <- Gest(ppp_csr) #<-Nearest Neighbor
f_csr <- Fest(ppp_csr) #<- for first-order distribution (non-random distribution)
k_csr <- Kest(ppp_csr) #<- for second-order distribution(clustering)

#Monte Carlo estimations
# This tests if the observed pattern is significantly different from random
g_csr_sim <- envelope(ppp_csr, Gest, nsim=99) #Monte-carlo estimations for CIs
f_csr_sim <- envelope(ppp_csr, Fest, nsim=99) #Monte-carlo estimations for CIs
k_csr_sim <- envelope(ppp_csr, Kest, nsim=99) #Monte-carlo estimations for CIs

#Plots side-by-side
par(mfrow = c(1, 3))

# CSR Plots
plot(g_csr_sim, main = "G-function M-C (CSR)")
plot(f_csr_sim, main = "F-function M-C (CSR)")
plot(k_csr_sim, main = "K-function M-C (CSR)")

mtext("CSR, as expected, is more or less randomly distributed,", side = 1, line = -3, outer = TRUE, cex = 1.2)
mtext("within CI of the expected lines and balanced, no first or second-order patterns", side = 1, line = -2, outer = TRUE, cex = 1.2)

```


```{r Question 2, part 2: Clustered}
# Clustered Pattern
g_cluster <- Gest(ppp_cluster)
f_cluster <- Fest(ppp_cluster)
k_cluster <- Kest(ppp_cluster)

# Monte Carlo Simulations
g_cluster_sim <- envelope(ppp_cluster, Gest, nsim=99)
f_cluster_sim <- envelope(ppp_cluster, Fest, nsim=99)
k_cluster_sim <- envelope(ppp_cluster, Kest, nsim=99)


#set up for side-by-sides
par(mfrow = c(1, 3))
# Clustered pattern Plots
plot(g_cluster_sim, main = "G-function (Clustered)") #<- nearest neighbor
plot(f_cluster_sim, main = "F-function (Clustered)") #<- first-order test (density of points across space)
plot(k_cluster_sim, main = "K-function (Clustered)") #<- second-order test (interactions among points, clustering)

mtext("G^ functions all have significantly higher cum. dist. of nearest-neighbors than random", side = 1, line = -3, outer = TRUE, cex = 1.2)
mtext("F-function shows significantly fewer points close to random points than random)", side = 1, line = -2, outer = TRUE, cex = 1.2)
mtext("K-function shows significantly more points closer to other points than random (clustered))", side = 1, line = -1, outer = TRUE, cex = 1.2)
```
```{r Question 2, part 3: segregated}
# Segregated Pattern
g_segregated <- Gest(ppp_segregated)
f_segregated <- Fest(ppp_segregated)
k_segregated <- Kest(ppp_segregated) #<-

# Monte Carlo Simulations
g_segregated_sim <- envelope(ppp_segregated, Gest, nsim=99)
f_segregated_sim <- envelope(ppp_segregated, Fest, nsim=99)
k_segregated_sim <- envelope(ppp_segregated, Kest, nsim=99)

#set up for side-by-sides
par(mfrow = c(1, 3))

# Segregated pattern Plots
plot(g_segregated_sim, main = "G-function (Segregated)")
plot(f_segregated_sim, main = "F-function (Segregated)")
plot(k_segregated_sim, main = "K-function (Segregated)")

mtext("G^ functions generally have a significantly lower nearest-neighbors than random", side = 1, line = -3, outer = TRUE, cex = 1.2)
mtext("F-function observations generally close to randomly expected)", side = 1, line = -2, outer = TRUE, cex = 1.2)
mtext("K-function shows a pattern not different from random (not clustered))", side = 1, line = -1, outer = TRUE, cex = 1.2)
```

3. Analyze the forest tree point pattern dataset from Harvard Forest
X• Load hf253-04-stems-2014.csv
X• Subset the data to (i) spatial extents of gx between c(600, 700) and gy between c(400, 500) and (ii)
retain only the four most abundant species in the original dataset.
x• Create a marked multivariate point pattern from the subsetted data, assigning the species code
as a qualitative mark. Note that you will need to convert the species code to a factor as follows:
factor(myData$speciesColumn, levels="vector of species").
• Use tools at your disposal, including plotting and statistical tests, to propose an interpretation of the
possible role of biotic processes, including intra- and interspecific interactions between the four tree
species, in generating the spatial pattern. For this latter task rely on the G-Function for statistical
analysis - as well as any others you found useful.
Hints

```{r Question 3: application to Harvard Forest, visual stuff first}

# Harvard forest CSV upload
hf_data <- read_csv("hf253-04-stems-2014.csv")

##Subset the data to (i) spatial extents of gx between c(600, 700) and gy between c(400, 500) and (ii) retain only the four most abundant species in the original dataset.

# Count the observances of each species
species_count <- hf_data %>%
  count(sp) %>%
  arrange(desc(n))  # Sort by abundance
#display top 4
head(species_count, n = 4) #<- top 4 1. tsugca, 2. acerru 3. betual, 4. kalmla

#subset to include top 4 species and associated g stuff
hf_subset <- hf_data %>%
  filter(gx >= 600 & gx <= 700, gy >= 400 & gy <= 500) %>%
  filter(sp %in% c("tsugca", "acerru", "betual", "kalmla"))

# Ensure species codes are factors with the correct levels
hf_subset$sp <- factor(hf_subset$sp, levels = c("tsugca", "acerru", "betual", "kalmla"))

# Define the observation window (based on gx and gy ranges)
win <- owin(xrange = c(600, 700), yrange = c(400, 500))

# Create the marked point pattern (ppp object)
ppp_hf_subset <- ppp(
  x = hf_subset$gx, 
  y = hf_subset$gy, 
  marks = hf_subset$sp,  # Species as the mark
  window = win
)

#make a multivariate ppp, with species marked
ppp_hf_subset <- ppp(hf_subset$gx, hf_subset$gy, window = win, marks = hf_subset$sp)

plot(ppp_hf_subset)
mtext("This is really busy, I'll try different visualizations", side = 1, line = -3, outer = TRUE, cex = 1.2)
```

```{r Question 3: visualizations}
# Create density raster for trees
density_image <- density(ppp_hf_subset)
plot(density_image, main = "All top 4 species combined")
mtext("There does seem to be a pattern, ", side = 1, line = -3, outer = TRUE, cex = 1.2)
mtext("looking at the distribution the N corners are low", side = 1, line = -2, outer = TRUE, cex = 1.2)

# Set up a plotting area with 1 row and 4 columns
par(mfrow = c(1, 4))
# each species
species_list <- levels(ppp_hf_subset$marks)

# mt list
density_list <- list()

# Loop through each species for density
for (species in species_list) {
  # point patterns by species
  ppp_subset <- ppp_hf_subset[ppp_hf_subset$marks == species]
  
  # density for subset
  density_image <- density(ppp_subset)
  
  # convert to data frame
  density_df <- as.data.frame(density_image)
  density_df$species <- species
  
  # store
  density_list[[species]] <- density_df
}

# Combine for plotting
combined_density <- bind_rows(density_list)

# Plotting densities by density
for (sp in species_list) {
  # by species
  species_data <- combined_density %>% filter(species == sp)
  
  # plot
  p <- ggplot(species_data) +
    geom_raster(aes(x = x, y = y, fill = value)) + #value auto-uses species
    scale_fill_viridis(name = "Density") +  
    coord_fixed() +  # Maintain aspect ratio
    labs(title = "Density of Species Occurrences",
       x = "X Coordinate (gx)",
       y = "Y Coordinate (gy)") +
  facet_wrap(~ species, ncol = 4) +  # 4 plots in a row
    theme_minimal() +
    theme(legend.position = "right")
  
  # Print out
  print(p)
  }

```
There are several patterns to observe here. Looking at the combined plot, there are higher densities of trees in the southern half, with a gradual lessening from the peak density in centered on the sw corner.
The Tsuga species mirrors the pattern of the overall trees, being most abundant in a band running from the western border, to the southern border of the plot. This could, perhaps, mean that this is the most abundant of the trees in count, maybe it's shrubby, small, and abundant while the others are more mature, and large.
The Acer species is most abundant in the southwest corner, and somewhat on the other side of the band of hemlocks.
The Belulua species is msot abundant in the most soutwestern portion and a bit on the central-southern portion of the plot.
Finally, the Kalmia species is most abundant in the southeastern portion of the plot.

```{r question 3 conitnued: statistical tests}
par(mfrow = c(1, 3))  # side-by-sides

# getting loopy
for (species in species_list) {
  # Subset the ppp object by species
  ppp_species <- ppp_hf_subset[marks(ppp_hf_subset) == species]
  
  # G-, F-, and K-function for the species
  g_species <- Gest(ppp_species)
  f_species <- Fest(ppp_species)
  k_species <- Kest(ppp_species)
  
  # Monte Carlo simulations for the species
  g_species_sim <- envelope(ppp_species, Gest, nsim = 99)
  f_species_sim <- envelope(ppp_species, Fest, nsim = 99)
  k_species_sim <- envelope(ppp_species, Kest, nsim = 99)
  
  # Plotting results for the species
  plot(g_species_sim, main = paste("G-function (", species, ")", sep = ""))
  plot(f_species_sim, main = paste("F-function (", species, ")", sep = ""))
  plot(k_species_sim, main = paste("K-function (", species, ")", sep = ""))
}
```
The G-function for each species is significantly higher than the expeced lines, indicating that we would reject the null hypothesis of complete spatial randomness (CSR) for all species. This would make sense unless there were some plot somewhere that had all things being equal along every point.

The F-function for each species is significantly lower than the expected line, this indicates that there are fewer points close to random points than would be expected if the distribution was random. This could be an indicater of a first-order distribution: the trees are not randomly spread about the plot. This makes sense because there will likely be different microhabitats as well as the constant competition for light, water, and other resources, as well as incorporating the difference in each that would be due to the varying terrain. The F-function could indicate some increased dispersion, this could come from either intra- or inter-species repulsion via competition.

The K-function for each species is significantly higher than the expected line, indicating the distribution is more clustered than random. This makes sense, since all of these are trees and this could limit seed dispersal. Also there may be more intra-specific interactions, perhaps through pupping or some mutual benefit to being in a copse of trees (sorry, I am a marine biologist, I may be barking up the wrong tree with these terms, but I decided to leaf them in there).

Overall, in all species we see that their dispersion deviates from what would be expected from a CSR distribution. The null hypothesis of random distance is rejected. Further tests show that there is some significant evidence of an outcome resembling repulsion within species, where there will be some distance between their conspecifics. Finally, the clustering evident from the K-function indicates that there is also likely some interspecific competition at play, since each species is more likely to be clustered within its own species, rather than interspersed with others despite the intraspecific repulsion that is evident.

